Exploring the Role of Feedback in Machine Learning Jupyter Notebooks

### 2024-07-15 report/emse24.tex
feat(report): emse24 updates

\lstdefinestyle{mystyle}{
  language=Python,
  basicstyle=\ttfamily\footnotesize,
  keywordstyle=\color{delim},
  stringstyle=\color{punct},
  commentstyle=\color{numb},
  breakatwhitespace=true,
  breaklines=true,
  keepspaces=true,
  showspaces=false,
  showstringspaces=false,
  showtabs=false,
  tabsize=2,
  stepnumber=1,
  numbersep=3pt,
  captionpos=b,
  % numbers=left,
  % numberstyle=\tiny\color{gray},
  % backgroundcolor=\color{backcolor},
  aboveskip=0pt,
  belowskip=0pt
}
\lstset{style=mystyle}

@@

\section{Introduction}

% TODO: feedback from Diomidis
% - In the introduction clarify (and perhaps illustrate) the life cycle and the role of notebooks.  Are these used to develop code that is then integrated into scripts, or is the notebook the execution vehicle, or is it informal throw-away code and rough notes on how to repeat a process in the future?  This will also inform the discussion and the implications.  Consider describing
%  a) the current development model, with the role of assertions, implicit checks, and visualization,
%  b) its limitations (which provide motivation for the study),
%  c) a sketch for an improved development model, and
%  d) the knowledge gaps associated with the improved model and how this study fills them.
%
% - The part starting "However, these decisions are often based on the conditions present when the model was initially created", which provides the motivation must be argued more convincingly.  I think that the motivation is not so much what is described in that paragraph, as the fact that, in the absence of detailed, up-to-date, and accurate requirements, rationale, and design documentation as well as tests, many of these can be derived from notebook interactions.

The machine learning development lifecycle (MLDL) encompasses a multifaceted series of entangled stages. Significant effort is invested in gathering and assembling the required data from diverse sources. The data then undergo cleaning and preprocessing to ensure high quality and usability, in the subsequent stages. MLDL is dominated by the model development phase which is highly experimental and iterative. This stage involves exploratory data analysis (EDA) to uncover underlying patterns, fitting the data to various models, and rigorously analyzing the results. Often, this stage may lead to additional rounds of feature engineering, where new attributes are created based on insights gained from initial models to enhance model performance. Practitioners may also engage in comparing and contrasting different models to identify the most effective approach, followed by fine-tuning selected models to optimize performance. The final model is then deployed into production where it is continuously monitored, and periodically retrained to adapt to new data or changing conditions~\cite{haakman2021ai,amershi2019software,sculley2015hidden}.

% The feedback-driven style of development in MLDL resembles principles of Agile software development and the Cross-Industry Standard Process for Data Mining (CRISP-DM). Agile methodologies emphasize continuous iteration of development and testing throughout the software lifecycle. In Agile, software is developed in incremental, rapid cycles resulting in small, incremental releases with each release building on previous functionality. Each iteration is reviewed and critiqued by the project team, which includes representatives of the project's various stakeholders.

The MLDL is fundamentally iterative, closely resembling Agile software engineering principles, which emphasize incorporating early feedback and working in small, manageable iterations~\cite{betz2018managing}. Similarly, CRISP-DM---the dominant process model in data mining---outlines a cyclical process that evolves from understanding the business objectives and data, data preparation, modeling, evaluation and finally deployment and monitoring~\cite{martinez-plumed2021crisp-dm}. Both systems advocate for a dynamic, iterative approach that enhances adaptability and effectiveness in dealing with complex, changing systems like those encountered in ML projects.

@@

Feedback is crucial for making important decisions throughout the MLDL. Practitioners rely on the output of cells to determine the appropriate data preprocessing techniques, detect erroneous assumptions, create new features to boost model performance, and select specific hyperparameters. However, these decisions are often based on the conditions present when the model was initially created. Visual outputs in particular, are open to interpretation~\cite{heer2010tour}. The viewer's background and expertise significantly influence how they understand and draw conclusions from visual data. 
For instance, interpreting a histogram requires knowledge of various distribution types, which someone without statistical training might not possess. Additionally, physical limitations such as color blindness can impede accurate interpretation if visualizations are not designed with accessibility in mind. As data, personnel, and system environments inevitably change, undocumented decisions can lead to accumulation of technical debt. This not only prolongs time required for debugging and troubleshooting, but also incurs additional costs and resource expenditures~\cite{sculley2015hidden,amershi2019software,sambasivan2021everyone}.

% TODO: feedback from Luis
% - not flowing with previous paragraph
% - The previous paragraph could mention something like ad-hoc manual tests. And then in this paragraph we could say that software engineering proposes practices on automated testing that can be useful in this context
% - in theory, the main goal of tests is to verify the requirements of the software. Documentation is a by product. (I like the idea in this paragraph but we dont want to touch a raw nerve from reviewers)
In software engineering, tests are employed to document the expected behaviour of code. Failing tests indicate when these expectations are no longer met. Similarly, assertions in Jupyter notebooks can serve as documentation for the decisions made during model development. Embedding assertions in the notebook can ensure that any deviations from expected data conditions or model performance are promptly detected, thus maintaining the integrity of the ML pipeline.

To understand the role of feedback when developing ML pipelines in Jupyter notebooks, we conduct a series of  empirical studies. First, we examine the outputs of code cells in Jupyter notebooks. Next, we study Python assertions written in Jupyter notebooks to understand how they are used when developing ML pipelines in notebooks. Finally, we investigate the intersection of outputs and assertions by analysing  assertions that document decisions made from the outputs of code cells. In particular, we focus on assertions derived from visualisations since they pose a higher risk of technical debt. To the best of our knowledge, we are the first to investigate how feedback derived from the output of code cells is used by ML practitioners when developing ML pipelines inside of Jupyter notebooks. We are also the first to investigate assertions written inside of notebooks and their intersection with the outputs of code cells.

The research questions along with the contributions of this study are presented below.

\begin{description}
  \item[RQ1.] \textbf{What are the characteristics of feebback mechanisms in ML Jupyter notebooks?}

  % TODO: the problem is that readers don't yet know that outputs are generated from last statement and print statements
  We mine 300K Jupyter notebooks written in Python from Github and Kaggle. We analyse 2.3M code cells and create a dataset of 22.5K assertions, 25K print statements and 28K last statements. We further perform a descriptive and lexical analysis of the dataset and report the key characteristics of feedback mechanisms used in ML Jupyter notebooks.

  % TODO: The dataset is available publicly on ...

  \item[RQ2.] \textbf{How is implicit feedback from the output of code cells used when writing ML code in Jupyter notebooks?}

  We conduct individual case-studies with XXX code cell outputs. During each case-study, we analyse the raw output along with the corresponding code that produced the output to understand its purpose. Additionally, we anayse the entire code cell, the previous and next cells along with the purpose of the notebook to bring in rich context.

  \item[RQ3.] \textbf{How are assertions used to validate ML code written in Jupyter notebooks?}

  Similarly, we conduct case-studies with XXX assert statements. We analyse the assertion along with the surrounding code to understand its purpose.

  \item[RQ4.] \textbf{How are assertions used to validate visual outputs in Jupyter notebooks?}

  Finally, we conduct case-studies with XXX visualisation and assertion (VA) pairs.
\end{description}

% TODO revisit this paragraph
Our findings indicate that outputs from code cells are extensively utilised for making critical design and implementation decisions during the MLDL. While manual validation of these outputs is common, it is also prone to inconsistencies thus highlighting the need for automated validation mechanisms. Assertions are identified as a potential solution, capable of verifying data distributions and ensuring model performance metrics meet predefined thresholds, thereby reducing reliance on manual checks and enhancing the robustness of ML pipelines. However, the study also notes the limited use of assertions, with only 9.7\% of the 3 million notebooks analyzed incorporating them, suggesting a significant gap in current testing practices within the ML community. The findings underscore the necessity for developing ML-specific testing tools and integrating assertions more deeply into the ML development workflow to ensure consistent and reliable validation processes.

@@

% TODO: feedback from Diomidis
% - In Section 2, rather than jumping to literate programming, it would be more appropriate to discuss the CLI interaction style, which also encompasses Jupyter.
% - Section 2.2 (Python AST) can be shortened, shortened and moved to methods, or removed. Consider also describing the format of Jupyter notebooks.

\subsection{Literate Programming, Computational Notebooks and Project Jupyter}

% NOTE pimentel2019large-scale
% NOTE kery2018story
% NOTE rule2018exploration

@@

Literate programming is a paradigm introduced by Donald Knuth that emphasizes writing code that is understandable by humans. In his seminal work, Knuth advocates combining source code with documentation in a single, cohesive narrative~\cite{knuth1984literate}. Computational notebooks are inspired by this paradigm, allowing for the integration of executable code, scientific equations, visualisations, and explanatory text within a single document. This format supports a literate approach to programming, making it easier to share and understand complex analyses and workflows. Project Jupyter is a prominent example of this approach, offering a suite of open-source tools for interactive computing. Jupyter Notebooks, in particular, have become a popular tool in the ML community since they facilitate open and reproducible science, education, and data analysis.

@@

\subsection{Outputs in Jupyter Notebooks}

@@

% explain the high-level way in which ast module in python works
% explain the anatomy of assert statements and the more complicated way in which asts for print statements are represented in an AST

An Abstract Syntax Tree (AST) is a tree representation of the abstract syntactic structure of source code. Each node in the tree represents a construct occurring in the source code. ASTs are widely used in compilers and code analysis tools to understand and manipulate code programmatically~\cite{wikipedia2024abstract}. The Python \emph{ast} module enables the parsing, manipulation, and analysis of Python source code by converting it into an AST. This module allows for the examination and modification of code structure, making it valuable for tasks such as code analysis, optimisation, and transformation. The ast module can parse source code strings into AST objects, traverse and modify the tree, and compile it back into executable code, providing a powerful tool for dynamic code manipulation~\cite{pythonast}.

In Python, \lstinline{assert} statements are represented in the AST by the \lstinline{Assert} class. The \lstinline{Assert} node includes the \lstinline{test} attribute---which contains the condition being asserted---and the \lstinline{msg} attribute---which holds an optional message displayed if the assertion fails. Print statements are treated as a call to the built-in \lstinline{print} function. Function calls are represented in the AST by the \lstinline{Call} class. The \lstinline{Call} node includes the \lstinline{func} attribute representing the \lstinline{print} function, and \lstinline{args} which lists the arguments passed.

@@

% TODO we need to explain why we only focus on the last statement of code cells (when visualisation code written anywhere produces an output); this is much harder to parse: we need to construct an exhaustive list of python visualisation libraries; people can use any naming convension to...

% The predominant form of obtaining feedback, while developing ML pipelines in Jupyter notebooks, is through the output of code cells. These outputs can come from three main sources in the source code--print statements, the last statement in code cells, and statements that create visualisations. Building on the principles from software engineering, we also recognize assert statements as a valuable form of feedback for writing ML code. This idea draws from prior literature in software engineering that connects test-driven development (TDD) as a form of developing software using feedback.

@@

Figure~\ref{fig:data-collection} summarises the data collection procedure used in this study. We collected Jupyter notebooks written in Python, from Kaggle and Github.

We mined public repositories from Github to collect $49923$ Jupyter notebooks. We use the Github advanced search syntax~\footnote{https://docs.github.com/en/search-github/searching-on-github/searching-code} to isolate Jupyter Notebook that contain the keyword ``assert'' in them. The final search query is as follows: \texttt{"assert" language:"Jupyter Notebook"}.\footnote{Collected on June 22, 2023}

For Kaggle, we used a pre-existing dataset KGTorrent~\cite{quaranta2021kgtorrent} since Kaggle does not support advanced code-based search like Github. To the best of our knowledge, KGTorrent this is the largest dataset of Python Jupyter notebooks obtained from Kaggle consisting of $248763$ notebooks.

Therefore, we start with 283GB of data comprising of $298686$ Python Jupyter notebooks.

% TODO: feedback from Luis
% - explain why we check the validity
% - we cannot guarantee that the white literature will be online in the next months/years and make it difficult to collect an important detail for reproducibility. I would add to the text at least how many libraries we added, give a few examples and point to our replication package,
We check the validity of the underlying JSON structure in each notebook using the nbformat tool~\footnote{https://nbformat.readthedocs.io/en/latest/}. Since the focus of this study is on the analysis of Python code, we only include notebooks that contain at least one code cell. Finally, to focus on Python code written specifically for ML projects, we only include notebooks that import popular ML libraries~\footnote{These libraries are derived from grey literature---https://www.coursera.org/articles/python-machine-learning-library, https://www.geeksforgeeks.org/best-python-libraries-for-machine-learning, https://www.datacamp.com/blog/top-python-libraries-for-data-science, https://www.kdnuggets.com/2020/11/top-python-libraries-data-science-data-visualization-machine-learning.html, https://lp.jetbrains.com/python-developers-survey-2022}.

For explicit feedback using assertions, we programmatically analyse the JSON structure of the notebooks to isolate code cells. We collect Python assert statements from the code cells by constructing and subsequently parsing the Abstract Syntax Tree (AST) of the source code present in each cell.

% TODO: feedback from Luis
% - only here, a reader start understanding that we want to collect assert, prints, and lasts. Perhaps we want to make that clear in the beginning of this subsection or in the one above it
Similarly, for implicit feedback, we first isolate code cells that produce an output. The output of a cell may either be from a ``print'' statement or the last statement of the cell. We collect both of them using the AST of the cell.

Finally, we remove duplicate data points (asserts, prints and last statements) resulting in a final data set of $22548$ assertions, $25319$ print statements and $28522$ last statements.

@@

% TODO: might need a visual aid here
% TODO: feedback from Luis
% - motivate this approach
% - not sure we want to say NLP here. I would call it text processing. NLP kinda implies that we are automatically interpreting the text
We allocated a fixed time resource of 100 hours to conduct the case study analysis of all candidates (assertions and cell outputs). To surface interesting candidates for the case studies, we apply NLP techniques as described below.

% TODO: point out the poor quality of notebooks out there (we have citations for this); and the resulting noise in the assertions/outputs. Thats why we need to apply these techniques to surface unique and interesting candidates!

The assertions and cell outputs are first tokenised---special characters and alpha-numeric words shorter than two characters are removed. Two stop words namely ``assert'' and ``print'' are removed since they appear in all assertions and print statements respectively. The term frequency (TF) for all tokens is calculated and then normalised using their inter-document frequency (IDF) such that tokens that appear less frequently are assigned a higher value.

We apply stratified random sampling to identify the candidates for the case study analysis. The sub-groups are created by adding TF-IDF of the tokens in each candidate to produce an aggregate value. The candidates are then divided into quartiles based on the aggregate value. A candidate is randomly drawn from each bin and analysed as an in-depth case study. The analysis is stopped when the time resource is exhausted.

During each case study, we analyse the code of the candidate to understand its purpose. Additionally, we analyse the entire code cell, the previous cell, next cell and the notebook's purpose to bring in rich context.

@@

% TODO: feedback from Diomidis
% - Sections 4.1 and 4.2 are very interesting but overly detailed. I suggest to move most of the content to an appendix or an online appendix, and only present the uses in a table or a taxonomy tree.  In addition, consider moving the more general observations into a discussion section.
% - Consider starting with RQ2 (making it RQ1), because assertions are a concrete feature of Python code.  Then move to RQ1, which deals with implicit assertions.
% - I am afraid I cannot understand RQ3; consider revising it.  I would think that visualizations are verified visually, rather than through assertions.  You may have encountered some parts, but I would assume that the assertions mostly deal with things that can be more accurately and effectively be expressed as a predicate rather than viewed in a chart.
% TODO: some sort of introduction to the following sub-sections here; we need to explain that for code cell outputs we try to show the python statement that produces the output along with the actual output. Somethimes actual output is not possible (eg. P68) because the statement may be inside an if statement.
\subsection{(RQ1) What are the characteristics of feedback mechanisms in ML Jupyter notebooks?}

@@

% TODO: some form of background on AST representation of assert statements is missing

@@

    \lstinline[]$torch.__version__ > '1.10.0'$&
    \lstinline[]$isinstance(transmat, pd.DataFrame)$\\

@@

We collected 1.4 million print statements from 180 thousand notebooks (60\% of the total notebooks analysed in this study). Unlike assert statements, print statements do not have a fixed lexical semantic since the Python ast module interprets them as regular function calls. Additionally, print statements can be written in several forms---the \lstinline{print} function may have multiple arguments, any form of string interpolation supported by Python or output from function calls. Figure~\ref{fig:common-print-nodes} shows that the most common nodes inside of print statements are constants followed by function calls. Figure~\ref{fig:common-print-constant-types} further validates that the constants are almost entirely of type string. Based on these observations, we disect the print statements into two components---text written in natural language (from the \emph{Constant} nodes) and code (from all other nodes). Figure~\ref{fig:common-print-constants} and Figure~\ref{fig:common-print-not-constants} present the most common bi-grams that appear in the natural text and code of print statements respectively. Both figures collectively indicate that print statements are frequently used to print various performance metrics of ML models.

@@

\subsection{(RQ2) How is feedback from the output of code cells used when writing ML code in Jupyter notebooks?}~\label{sec:result-rq1}
% TODO: feedback from Luis
% -  up to this point I miss how exactly we plan to answer this question. Reading the results I get that we are answering by clustering outputs together based on their respective ML/DS task and then by analysing how that task is accomplished. We need to explain this either in the beginning of section 4.1 or in section 3

>>>>

\lstdefinestyle{mystyle}{
  language=Python,
  basicstyle=\ttfamily\footnotesize,
  keywordstyle=\color{delim},
  stringstyle=\color{punct},
  commentstyle=\color{numb},
  breakatwhitespace=true,
  breaklines=true,
  keepspaces=true,
  showspaces=false,
  showstringspaces=false,
  showtabs=false,
  tabsize=2,
  stepnumber=1,
  numbersep=3pt,
  captionpos=b,
  % numbers=left,
  % numberstyle=\tiny\color{gray},
  backgroundcolor=\color{backcolor},
  % aboveskip=0pt,
  % belowskip=0pt
}
\lstset{style=mystyle}

@@

\section{Introduction}
The machine learning development lifecycle (MLDL) encompasses a multifaceted series of entangled stages. Significant effort is invested in gathering and assembling the required data from diverse sources. The data then undergo cleaning and preprocessing to ensure high quality and usability, in the subsequent stages. MLDL is dominated by the model development phase which is highly experimental and iterative. This stage involves exploratory data analysis (EDA) to uncover underlying patterns, fitting the data to various models, and rigorously analysing the results. Often, this stage may lead to additional rounds of feature engineering, where new attributes are created based on insights gained from initial models to enhance model performance. Practitioners may also engage in comparing and contrasting different models to identify the most effective approach, followed by fine-tuning selected models to optimize performance. The final model is then deployed into production where it is continuously monitored, and periodically retrained to adapt to new data or changing conditions~\cite{haakman2021ai,amershi2019software,sculley2015hidden}.

The MLDL is fundamentally iterative, closely resembling Agile software engineering principles, which emphasize incorporating early feedback and working in small, manageable iterations~\cite{betz2018managing}. Similarly, CRISP-DM---the dominant process model in data mining---outlines a cyclical process that evolves from understanding the business objectives and data, data preparation, modelling, evaluation and finally deployment and monitoring~\cite{martinez-plumed2021crisp-dm}. Both systems advocate for a dynamic, iterative approach that enhances adaptability and effectiveness in dealing with complex, changing systems like those encountered in ML projects.

@@

% NOTE: consider starting with prints and last statements first (most obvious), then move to assertions; its fine if RQs follow different order
Jupyter notebooks are primarily used to explore different ideas or approaches, often neglecting software development best practices in favor of rapid experimentation~\cite{kery2018story,rule2018exploration,pimentel2019large-scale}. During this process, practitioners can use various feedback mechanisms like assertions, print statements, output of code cells and visualisations to determine the appropriate data preprocessing techniques, detect erroneous assumptions, create new features to boost model performance, and select specific hyperparameters~\cite{head2019managing,chattopadhyay2020whats}. Once the code is finalised, it is typically transferred into a script that is run in a production environment. Alternatively, the notebook may contain significant exploratory work such as developing a model or performing data exploration. To effectively communicate the information present in the notebooks with clients, collaborators or teammates, the practitioners add annotations and structure to the notebook using markdown cells~\cite{kery2018story,rule2018exploration}. In both scenarios, practitioners need to revisit the notebooks to obtain structured information such as up-to-date and accurate requirements, their rationale and design decisions, and tests---artifacts which are usually missing in the original notebooks~\cite{pimentel2019large-scale,psallidas2019data,grotov2022large-scale}. As data, personnel, and system environments inevitably change, undocumented decisions can lead to accumulation of technical debt. This not only prolongs time required for debugging and troubleshooting, but also incurs additional costs and resource expenditures~\cite{sculley2015hidden,amershi2019software,sambasivan2021everyone}.

% TODO: feedback from diomidis
% - "We identify a knowledge gap in the current approach used to obtain such structured information from experimental code within Jupyter notebooks." Rephrase to have the knowledge gap correspond to what the RQs answer.  Currently one may think that we aim to improve the approach, not its understanding.
We identify a knowledge gap in the current approach used to obtain such structured information from experimental code within Jupyter notebooks. To address this gap, this study conducts a series of empirical studies to understand the role of various feedback mechanisms used to develop ML pipelines within Jupyter notebooks. We focus on implicit form of feedback obtained from print statements and output generated by the last statement in code cells, as well as explicit form of feedback obtained from assertions. To the best of our knowledge, we are the first to investigate how such feedback mechanisms are used by ML practitioners when developing ML pipelines inside Jupyter notebooks. The research questions along with the contributions of this study are presented below.

\begin{description}
  \item[RQ1.] \textbf{What feedback mechanisms are employed in the development of ML Jupyter notebooks?}

    We mine 297.8 thousand Jupyter notebooks written in Python from Github and Kaggle. We analyse 2.3 million code cells obtained from the notebooks and create a public dataset of 27.1 thousand assertions, 498.3 thousand print statements and 373.6 thousand last statements. We perform a descriptive and lexical analysis of the dataset and report the key characteristics of feedback mechanisms used in ML Jupyter notebooks.

  % TODO: add citation for the dataset (make it public so that DOI is available)

  \item[RQ2.] \textbf{How is explicit feedback from assert statements used to validate ML code written in Jupyter notebooks?}

    To gain a deeper understanding of assert statements written in Jupter notebooks, we conduct individual case-studies with 82 assert statements. We analyse the assertion along with the surrounding code to understand its purpose. Additionally, we analyse the entire code cell, the previous and next cells along with the purpose of the notebook to bring in rich context.

  \item[RQ3.] \textbf{How is implicit feedback from print statements and last statements of code cells used when writing ML code in Jupyter notebooks?}

    Similarly, we conduct individual case-studies with 44 print statements and 27 last statements. During each case-study, we analyse the code statement along with the raw output to understand its purpose.
\end{description}

% TODO revisit this paragraph
% TODO what should be done to change the things we observe? to increase the use of explicit feedback? do we know if its important or not? maybe we need to explore that?
Our findings indicate that outputs from code cells are extensively utilised for making critical design and implementation decisions during the MLDL. While manual validation of these outputs is common, it is also prone to inconsistencies thus highlighting the need for automated validation mechanisms. Assertions are identified as a potential solution, capable of verifying data distributions and ensuring model performance metrics meet predefined thresholds, thereby reducing reliance on manual checks and enhancing the robustness of ML pipelines. However, the study also notes the limited use of assertions, with only 9.7\% of the 3 million notebooks analyzed incorporating them, suggesting a significant gap in current testing practices within the ML community. The findings underscore the necessity for developing ML-specific testing tools and integrating assertions more deeply into the ML development workflow to ensure consistent and reliable validation processes.

@@

\subsection{REPLs, Interactive Computing and Project Jupyter}

A Read-Eval-Print Loop (REPL) is an interactive programming environment that takes single user inputs (Read), executes them (Eval), returns the result to the user (Print) and then waits for the next command (Loop). REPLs offer an immediate feedback loop for quickly testing and debugging code snippets. The Python standard distribution includes a Python REPL with a command line interface to execute Python code in real-time. IPython---a more advanced alternative to the Python REPL---extends the capabilities of the standard Python shell with additional features such as powerful introspection, additional shell syntax, tab completion, and command history. Jupyter notebooks (and its successor Jupyter lab) leverage the IPython kernel to execute Python code and extends its capabilities with a graphical user interface (GUI). Resembling the literate programming paradigm~\cite{knuth1984literate}, Jupyter notebooks offer a web-based interface where users can create and share documents that contain live code, equations, visualisations, and narrative text. This format supports a literate approach to programming, making it easier to share and understand complex analyses and workflows. Jupyter Notebooks in particular, have become a popular tool in the ML community since they facilitate open and reproducible science, education, and data analysis.

\subsection{Anatomy of Jupyter Notebooks}

@@

@@

@@

The Python \lstinline{ast} module enables the parsing, manipulation, and analysis of Python source code by converting it into an abstract syntax tree (AST)~\cite{pythonast}. The \lstinline{ast} module can parse source code strings into AST objects, traverse and modify the tree, and compile it back into executable code, providing a powerful tool for dynamic code manipulation. In Python, \lstinline{assert} statements are represented in the AST by the \lstinline{Assert} class. The \lstinline{Assert} node includes the \lstinline{test} attribute---which contains the condition being asserted---and the \lstinline{msg} attribute---which holds an optional message displayed if the assertion fails. Print statements are treated as a call to the built-in \lstinline{print} function. Function calls are represented in the AST by the \lstinline{Call} class. The \lstinline{Call} node includes the \lstinline{func} attribute representing the \lstinline{print} function, and \lstinline{args} which lists the arguments passed.

@@

@@

% TODO: I suspect this will raise eyebrows because the notebooks collected from Github were for the ICST paper, where we were also looking for the keyword "assert" (which does not apply to this paper)

Figure~\ref{fig:data-collection} illustrates the data collection process employed in this study to gather Jupyter notebooks written in Python from GitHub and Kaggle. We used Github's advanced search syntax~\footnote{https://docs.github.com/en/search-github/searching-on-github/searching-code}, with the following search query: \lstinline[language={}]$language:"Jupyter Notebook"$ to mine 49.9 thousand Jupyter notebooks from public repositories on Github~\footnote{Collected on June 22, 2023}. We used the pre-existing dataset KGTorrent~\cite{quaranta2021kgtorrent} for Kaggle since it does not support advanced code-based search like GitHub. To the best of our knowledge, KGTorrent is the largest dataset of Python Jupyter notebooks obtained from Kaggle, consisting of 247.9 thousand notebooks. Consequently, we started with 283GB of data comprising 297.8 thousand Python Jupyter notebooks.

Since the focus of this study is on the analysis of Python code, we only include notebooks that have a valid underlying JSON structure and contain at least one code cell. Finally, to focus on Python code written specifically for ML projects, we only include notebooks that import at least one of the following popular ML libraries which we identified from grey literature~\cite{datacampml,gfgml,jetbrainsml,kdnuggetml,courseraml}---Scikit Learn~\cite{virtanen2020scipy}, Pytorch~\cite{paszke2017automatic}, Tensorflow~\cite{abadi2015tensorflow} and Keras~\cite{chollet2015keras}. We use the Python \lstinline{ast} module to extract the assert, print and last statements from all code cells. Prior to conducting the analysis for RQ2 and RQ3, we remove duplicate data points resulting in a final data set of 27.1 thousand assertions, 498.3 thousand print statements and 373.6 thousand last statements.

@@

We allocated a fixed time resource of 200 hours to conduct the case study analysis of all candidate assertions, print statements and last statements. To surface interesting candidates for the case studies, we apply text processing techniques as described below.

The assertions and cell outputs are first tokenised---special characters and alpha-numeric words shorter than two characters are removed. Two stop words namely ``assert'' and ``print'' are removed since they appear in all assertions and print statements respectively. The term frequency (TF) for all tokens is calculated and then normalised using their inter-document frequency (IDF) such that tokens that appear less frequently are assigned a higher value. We apply stratified random sampling to identify the candidates for the case study analysis. The sub-groups are created by adding TF-IDF of the tokens in each candidate to produce an aggregate value. The candidates are then divided into quartiles based on the aggregate value. A candidate is randomly drawn from each bin and analysed as an in-depth case study. The analysis is stopped when the time resource is exhausted. During each case study, we analyse the code of the candidate to understand its purpose. Additionally, we analyse the entire code cell, the previous cell, next cell and the notebook's purpose to bring in rich context.

@@

% TODO: feedback from Diomidis
% - Sections 4.1 and 4.2 are very interesting but overly detailed. I suggest to move most of the content to an appendix or an online appendix, and only present the uses in a table or a taxonomy tree.
% TODO: In addition, consider moving the more general observations into a discussion section.

\begin{figure}
  \centering
  \includegraphics[width=0.75\textwidth]{taxonomy.pdf}
  \caption{Overview of various feedback mechanism groups identified in this study.}
  \label{fig:taxonomy}
\end{figure}

% TODO: add citation to online appendix
The results of the study are presented in this section. For the case-studies conducted for RQ2 and RQ3, we group the candidates based on their similarity to each other. Please note that this is done only to facilitate structuring the paper in a meaningful manner and for easy of reporting the results. Figure~\ref{fig:taxonomy} provides an overview of the various feedback mechanism groups identified in this study. Additional details are provided in Section~\ref{sec:result-explicit} and Section~\ref{sec:result-implicit}. Throughout the text, we reference the candidates (\textbf{(A)}ssertions, \textbf{(P)}rint statements and \textbf{(L)}ast cell statements) using their unique identifiers as examples. These can be viewed on our online appendix.

\subsection{(RQ1) What feedback mechanisms are employed in the development of ML Jupyter notebooks?}~\label{sec:result-analysis}

@@

@@

    \lstinline[]$torch.__version__ > '1.10.0'$&
    \lstinline[]$isinstance(transmat, pd.dataframe)$\\

@@

We collected 1.4 million print statements from 180 thousand notebooks (60\% of the total notebooks analysed in this study). Unlike assert statements, print statements do not have a fixed lexical semantic since the Python \lstinline{ast} module interprets them as regular function calls. Additionally, print statements can be written in several forms---the \lstinline{print} function may have multiple arguments, any form of string interpolation supported by Python or output from function calls. Figure~\ref{fig:common-print-nodes} shows that the most common AST nodes inside of print statements are constants followed by function calls. Figure~\ref{fig:common-print-constant-types} further validates that the constants are almost entirely of type string. Based on these observations, we disect the print statements into two components---text written in natural language (from the \emph{Constant} nodes) and code (from all other nodes). Figure~\ref{fig:common-print-constants} and Figure~\ref{fig:common-print-not-constants} present the most common bi-grams that appear in the natural text and code of print statements respectively. Both figures collectively indicate that print statements are frequently used to print various performance metrics of ML models.

@@

\subsection{(RQ2) How is explicit feedback from assert statements used to validate ML code written in Jupyter notebooks?}~\label{sec:result-explicit}

Data shape checks can be considered the ``Swiss army knife'' of assertions, as they are ubiquitous and serve multiple purposes throughout the MLDL. We encountered a vast array of assert statements that verify the dimensions of various data structures, including input features (\emph{A17, A90}), labels (\emph{A5, A29}), predictions, images (\emph{A76, A93}), sequences, and embeddings (\emph{A76, A93}). These assertions ensure that the data adheres to the expected format and dimensions required by different components of the machine learning pipeline.

\highlight{\textbf{Finding 2.1} 26 assertions verify the dimensions of various data structures such as input features, labels, predictions, images, sequences, and text embeddings. These checks ensure data consistency and compatibility throughout the ML pipeline, preventing runtime errors and ensuring accurate model training and evaluation.}

We found assert statements that perform various validation checks on data structures, such as NumPy arrays and Pandas dataframes. These checks help ensure the data meets specific criteria or constraints, preventing potential errors or inconsistencies in downstream operations.

Common data validation checks include verifying the presence of specific values or ranges within arrays or columns(\emph{A41, A65}) and verifying the equality or closeness of arrays to a target value within a specified absolute tolerance (\emph{A44, A46}). Some assertions also focus on validating the uniqueness or cardinality of values (\emph{A52}) while others validate the presence of specific values or conditions within data structures (\emph{A73}).

\highlight{\textbf{Finding 2.2} Fourteen assertions perform validation checks on data structures, ensuring the presence of specific values, verifying value ranges, and confirming uniqueness or cardinality of values. These checks catch potential errors or inconsistencies early, ensuring data meets specific criteria or constraints for robust ML applications.}

\begin{lstlisting}[caption={Assertion \emph{A58} used to check that the accuracy of a model is close to the specified value. The use of \lstinline{np.isclose} allows for small deviations in the accuracy thus accounting for the stocastic nature of ML models.}, label={lst:A58}]
assert np.isclose(accuracy, 0.9666666666666667)
\end{lstlisting}

This study finds several assertions used to test model performance against predefined thresholds for key metrics such as accuracy, precision, recall, F1 score, and others. Range checks confirm that the performance metrics fall within acceptable boundaries, indicating good predictive performance without overfitting (\emph{A38}). Direct comparisons of model outputs with expected results, validate the model's predictive accuracy and reliability under operational conditions (\emph{A15}). Other assertions ensure that the learned parameters align closely with theoretically or empirically derived values, further cementing the model’s statistical validity (\emph{A19}). Finally assertions are also used to check for the expected number of neighbours as an indirect measure of model performance in specifc scenarios like clustering (\emph{A7}).

The use of \lstinline{np.isclose} in Listing~\ref{lst:A58} is particularly noteworthy. This method accounts for the stochastic nature of many ML models, where slight variations in performance metrics can occur due to differences in initial conditions, random seed settings, or inherent randomness in algorithms. By allowing a small tolerance in the comparison, \lstinline{np.isclose} ensures that the model's performance is consistently close to the expected benchmark, thereby affirming its reliability despite the stochastic elements.

\highlight{\textbf{Finding 2.3} Eleven assertions evaluate model performance against predefined thresholds for accuracy, precision, recall, F1 score, and other metrics. These checks validate the model's predictive accuracy and reliability, ensuring it meets expected performance standards despite the stochastic nature of ML models.}

Existence checks are primarily used during the data preprocessing stage in the MLDL to ensure data integrity before analysis and modelling. These checks primarily focus on verifying the presence of necessary columns and the absence of missing values within those columns (\emph{A23, A42, A79, A86}). Such validations are especially crucial after data preprocessing steps, where transformations might inadvertently introduce \lstinline{NaN} values (\emph{A50, A51, A63}) or result in an empty dataframe (\emph{A43}). These checks are also crucial for maintaining the accuracy and efficiency of data handling processes, ensuring that the datasets are ready for robust ML applications.

\highlight{\textbf{Finding 2.4} Eight assertions ensure the presence of specific columns in the dataset and the absence of missing values. These checks prevent operations on empty datasets and confirm the absence of \texttt{NaN} values, maintaining data integrity and reliability for model training.}

\subsubsection{Resource check ($N = 7$)}~\label{sec:result-rq2-resource-check}
% (A10, A14, A18, A37, A60, A67 A74)

\begin{lstlisting}[caption={Assertion \emph{A37} used to ensure that an ML model has not reached an inconsistent state due to out-of-order or re-execution of code cells.}, label={lst:A37}]
assert svm.fit_status_ == 0, 'Forgot to train the SVM!'
\end{lstlisting}

Resource check assertions are used to ensure the availability and validity of essential resources, preventing runtime errors, and maintaining the integrity of the data, models, and visualisations. One set of assertions verify the existence of files on the file system, such as pre-trained models (\emph{A10}) or data files (\emph{A74}). Subsequently, assertions are employed to validate the successful loading of models (\emph{A14}) and the completion of model training processes (\emph{A37}).

A unique aspect of working with notebooks is the ability to re-run cells while experimenting, which can lead to unintended consequences. Aassertion \emph{A37} presented in Listing~\ref{lst:A37} addresses this scenario by checking if the SVM model has been properly trained before proceeding with further operations. This check prevents inconsistent model states from executing cells out of order or multiple times. Other interesting assertions unique to computational notebooks include verifying the presence of data in visualisations (\emph{A60}) and explicitly verifying the versions of critical libraries (\emph{A18, A67}).

\highlight{\textbf{Finding 2.5} Seven assertions confirm the availability of essential resources like pre-trained models, data files, successful model loading, and completion of training processes. These checks prevent runtime errors and ensure proper code execution, maintaining the reliability and robustness of the ML workflows.}
